Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-02-27 Publications/2015-NeuroImage/03-introduction.tex
paper version candidate

  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, align geometrical features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} map geometrical features and voxel-wise information.

@@

\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe  in many different orientations the diffusion process within the living
  brain, these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwith originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations, $\Delta B_0$, is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bones in both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

@@

where $\vec{r}$ is the location in the reference domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution was acquiring extra images to estimate $\Delta B_0$ using field mapping
  techniques \citep{jezzard_correction_1995,andersson_modeling_2001}.
A second set of methods aquired extra images reproducing the same \gls*{epi} scheme but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient
  increments in the same axis \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \gls*{dmri}
  data versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
All methods require extra \gls*{mr} aquisitions. Thus, a tipical protocol for the extraction
  of structural connectivity comprehends \gls*{dmri}, \gls*{t1}, and the acquisition demanded
  by the correction method of choice.

Our method is related to the \gls*{t2b} methods as we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
Similarly, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}),
  and then map these surfaces into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

>>>>

  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classified the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are referred to as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, align spatial features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} map geometrical features and voxel-wise information.

@@

\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe the diffusion process within the brain in many different orientations,
  these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwith originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations ($\Delta B_0$) is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bone of both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

@@

where $\vec{r}$ is the true location in the undistorted domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution was acquiring extra images to estimate $\Delta B_0$ using field mapping
  techniques \citep{jezzard_correction_1995,andersson_modeling_2001}.
A second set of methods aquired extra images reproducing the same \gls*{epi} scheme but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient
  increments in the same axis \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \gls*{dmri}
  data versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
All methods require extra \gls*{mr} aquisitions. Thus, a tipical protocol for the extraction
  of structural connectivity comprehends \gls*{dmri}, \gls*{t1}, and the acquisition demanded
  by the correction method of choice.

\emph{Regseg} is related to the methods based on nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value used as signal reference.}
  to the \gls*{t2} image, since we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align contours extracted from the \gls*{t1}
  to \gls*{dmri} images using a linear transformation model.
Similarly, we exploit the precise surfaces that can be extracted with
  \emph{FreeSurfer}~\citep{fischl_freesurfer_2012},
  and then map this prior into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
\cite{greve_accurate_2009} circumvented the nonlinearity by identifying tipically
  distorted regions and removing them from the metric computation.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving (distorted) coordinate system.

### 2015-02-27 Publications/2015-NeuroImage/04-methods.tex
paper version candidate

% -*- root: 00-main.tex -*-

%{\color{red} {The main diverging points with respect to
%\citep{gorthi_active_2011} are: 1) there is no need for an explicit level set function
%$\Phi_G$, as we replace the level set gradient computation $N_{\Phi_G}$ with shape
%gradients \citep{besson_dream2s_2003,herbulot_segmentation_2006};
%2) regularization is also based on linear diffusion smoothing \citep{thirion_image_1998},
%but we replace the Gaussian filtering by other constraints
%studied in \citep{nagel_investigation_1986} to better the problem;
%3) optimization is applied in the spectral
%domain, observing anisotropic and inhomogeneous mappings along each direction.
%With respect \citep{guyader_combined_2011}, the main differences are
%the distance function, and the spectral solution to the optimization updates,
%as we shall cover in \autoref{sec:numerical_implementation}.}}

%In this paper we formulate the joint registration-segmentation
%problem as follows.
%such that the known contours in anatomical space $T$ optimally segment
%the diffusion space $D$.
%
%
%Whereas related \glspl*{adf} introduced in \autoref{sec:methods_background}
%make use of explicit level-set formulations to solve \eqref{eq:gradient_descent},
%we alternatively use \emph{shape-gradients}
%\cite{besson_dream2s_2003,herbulot_segmentation_2006}.

%
%Let us denote by $\vec{x}$ the voxel and $F(\vec{x}) = [ f_1, f_2, \ldots, f_N]^T$
%  its associated feature vector in the following.
%
%In our application, these surfaces are precise tissue interfaces of interest extracted
%  from a high-resolution, anatomically correct reference volume using
%  the well-established
%
%In this paper we propose a novel registration framework to simultaneously
%solving the segmentation, distortion and cortical parcellation challenges,
%by exploiting as strong shape-prior the detailed morphology extracted
%from high-resolution and anatomically correct \gls{mri}.
%Indeed, hereafter
%we assume this segmentation problem in anatomical images is reliably and
%accurately solved with readily available tools (e.g.
%\citep{fischl_freesurfer_2012}).
%After global alignment with \gls{t1} using existing approaches, the remaining
%spatial mismatch between anatomical and diffusion space is due to susceptibility
%distortions.
%Finally, we need to establish precise spatial correspondence between the
%surfaces in both spaces, including the tangential direction for parcellation.
%Therefore, we can reduce the problem to finding the differences of spatial
%distortion in between anatomical and \gls{dwi} space.
%We thus reformulate the segmentation problem as an inverse problem, where we
%seek for an underlying deformation field (the distortion) mapping
%from the structural space into the diffusion space, such that the structural
%contours segment optimally the \gls{dwi} data.
%In the process, the one-to-one
%correspondence between the contours in both spaces is guaranteed, and projection
%of parcellisation to \gls{dwi} space is implicit and consistent.
%
%We test our proposed joint segmentation-registration model on two different
%synthetic examples.
%The first example is a scalar sulcus model, where the
%\gls{csf}-\gls{gm} boundary particularly suffers from \gls{pve} and can only be
%segmented correctly thanks to the shape prior and its coupling with the inner,
%\gls{gm}-\gls{wm} boundary through the imposed deformation field regularity.
%The second case deals with more realistic \gls{dwi} data stemming from
%phantom simulations of a simplistic brain data.
%Again, we show that the
%proposed model successfully segments the \gls{dwi} data based on two derived
%scalar features, namely \gls{fa} and \gls{md}, while establishing an estimate
%of the dense distortion field.
%
%The rest of this paper is organized as follows.
%First, in \autoref{sec:methods}
%we introduce our proposed model for joint multivariate segmentation-registration.
%Then we provide a more detailed description of the data and experimental setup in
%\autoref{sec:experiments}.
%We present results in \autoref{sec:results} and conclude
%in \autoref{sec:conclusion}.
%
%
%The properties of the reconstructed tensors and derived scalar maps have
%been studied by \cite{ennis_orthogonal_2006}. Based on their
%findings, \gls{fa}~\eqref{eq:fa} and \gls{md}~\eqref{eq:md} are
%considered complementary features, and therefore we selected them for the
%energy model \eqref{eq:complete_energy} in driving the
%registration-segmentation process.
%Whereas \gls{fa} informs mainly about the \emph{shape} of diffusion,
%the \gls{md} is more related to the \emph{magnitude} of the process:
%
%\begin{align}
%\mathrm{FA} &= \sqrt{ \frac{1}{2}}\,\frac{\sqrt{ (\lambda_1 - \lambda_2)^2 + (\lambda_2 - \lambda_3)^2 + (\lambda_3 - \lambda_1)^2}}{\sqrt{ {\lambda_1}^2 + {\lambda_2}^2 + {\lambda_3}^2}} \label{eq:fa} \\
%\mathrm{MD} &= ( \lambda_1 + \lambda_2 + \lambda_3 ) / 3 \label{eq:md}
%\end{align}
%where $\lambda_i$ are the eigenvalues of the diffusion tensor
%associated with the diffusion signal $S(\vec{q})$. There exist
%two main reasons to justify their choice.
%First, they are well-understood and standardized in clinical routine.
%Second, together they contain most of the information that is
%usually extracted from the \gls{dwi}-derived scalar maps
%\cite{ennis_orthogonal_2006}.
%

\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built including a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well
  \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.
In \autoref{sec:methods_map} we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level sets
  framework and only one \gls*{pde} for first.
The principal application of these techniques was the image registration between timesteps in
  a time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on the level sets approximation in the application of
  atlas-based segmentation.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method close to the one proposed here.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the smoothing effects of the deformation field and the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

@@

  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mathcal{E}_l(\vec{r}) = \mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.

@@

  \begin{align*}
  P(U) = \underset{\vec{r}}{\prod}\, p(u(\vec{r})) &=
  \underset{\vec{r}}{\prod}\, p_0(u(\vec{r})) \, p_1(u(\vec{r})), \\
  p_0(u(\vec{r})) &= \mathcal{N}( u(\vec{r}) \mid 0, A^{-1}), \notag\\
  p_1(u(\vec{r})) &= \mathcal{N}(  \nabla \cdot u(\vec{r}) \mid 0, B^{-1}),
  \end{align*}

  imposing that the distortion and its gradient have zero
  mean and variance governed by $A$ and $B$.
Since the anisotropy is generally aligned with the imaging axes, these will be simplified
  in the following for the sake of clarity:

@@

Finally, the \gls{map} problem is adapted into a variational one applying the
  following log-transform:

@@

  that is the dual expression to the energy functional corresponding
  to a discrete \gls*{acwe} framework \citep{chan_active_2001}
  with anisotropic regularization as studied in
  \citep{nagel_investigation_1986}.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the segmenting
	\glspl{roi} $\Omega_i$, and represented in green and dark blue colors in this
	close-up.
	They iteratively through their normals $\hat{n}_i$ at each vertex $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	the vertex $\vec{v}_i$ in each limiting region as described in \eqref{eq:energy}.
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer.
	Finally, every $\bar{s}_i$ is projected into the grid of control points $\vec{u}_k$ that
	support the deformation field through the corresponding weights $\psi_k(\vec{v}_i)$ obtained
	as in \eqref{eq:basis_derivative}.
	}\label{fig:method}
\end{figure}

@@

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
  to compute the data term in \eqref{eq:gradient_descent} as follows:

@@

  in this case, the formulation has been adapted to the non-binary case, $\{l,m\}$
  being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
  them such that $\vec{x}' = \vec{v}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$
  and $\vec{n_i}'$ is the unit inward normal to the contour at $c_i'$.

@@

\paragraph*{Settings, implementation details, and convergence}
\label{sec:conv_report}
The registration parameters (such as $\tau$, the B-Spline grid resolutions,
  target image smoothing, etc.) and implementation details (for instance the sparse matrix approach
  to fast interpolation) are dicussed in the \suppl{}, {\color{red} section S1}.
The actual choice of parameter settings are publicly distributed with the source code of the experiments.
Additionally, we release our software along with a tool for generating convergence reports to
  demonstrate the behavior of the tool and help scientists configure their experiments.
A sample of report is found in the \suppl{} {\color{red}, section S1.3}.

\subsection*{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
In order to demonstrate the applicability of the registration method, we first
  subjected it to a battery of random tests over synthetic data and then applied
  it on real data.
\autoref{fig:evworkflows} presents the workflow implementing the evaluation instruments.
Besides a visual assessment of the results, we report quantitative evaluations using
  two metrics.
In the case of the phantoms, as we warped data along the three available dimensions,
  we computed the Haussdorf distance, integrating the ``point-to-cell'' method
	implemented by \cite{commandeur_vtk_2011}.
Conversely, the susceptibility-derived distortions only happen along the phase
  encoding axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $w_i$:

  \begin{equation}
  sWI = \frac{1}{P} \sum\limits_p^P \frac{1}{A_p} \sum\limits_i^{N_p} w_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \end{equation}

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figures/figure02}
\caption{Experimental workflow applied on real data from the \gls*{hcp}.
  1) The prior surfaces are extracted from the anatomical reference (\gls*{t1} image).
	2) To operate as ground truth, we generate plausible-synthetic distortion $U^{-1}_{true}$
	  from the fieldmap using \eqref{eq:fieldmap}.
	3) The \gls*{dmri} data are warped using $U^{-1}_{true}$ to reproduce the effects of real
	  susceptibility-derived distortions.
	Target diffusion scalars (\gls*{fa} and \gls*{adc}) are computed on the distorted data and
		stacked to feed the multivariate input required by our algorithm.
	4) Registration is run, obtaining a $U_{test} = \hat{U}_{true}$, an estimation of
	  the ground-truth deformation $U_{true}$.
	  A cross-comparison methodology is also applied, to obtain a competing $U_{cc}$.
	5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure*}

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each appearance shape, with two resolutions each
  ($1.0mm$ and $2.0mm$ isotropic).
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted using the marching-cubes algorithm
  shipped with \emph{FreeSurfer} \citep{fischl_freesurfer_2012}
  at the higher resolution (1.0mm isotropic).

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing with
  \emph{FreeSurfer}.

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures;
  3) cerebral \gls*{wm};
  4) brain stem and cerebellar \gls*{wm};
	5) cerebellar \gls*{gm}; and
	6) cortical \gls*{gm} surface.
The choice of this particular model is further addressed in \nameref{sec:res_model_and_metric}.

In this case, we derived the deformation $U_{true}^{-1}$ from the field maps released with
  the corresponding packages of each dataset from the \gls*{hcp} applying \eqref{eq:fieldmap},
  mimicking the real distortions, using a derivation of our previous work
  \citep{esteban_simulationbased_2014}.
After warping the original \gls*{dmri} with $U_{true}^{-1}$, we computed the \gls*{dti} and
  the derived scalars (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.

>>>>

% -*- root: 00-main.tex -*-
\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built on a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well
  \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.
In section \nameref{sec:methods_map}, we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	two existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level-sets
  framework and only one \gls*{pde} for first time.
The principal application of these techniques was to perform registration between timesteps in
  time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on level sets in the application of atlas-based segmentation.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method close to the one proposed here.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the inherent smoothness of the deformation field, and on the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

@@

  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.

@@

  \begin{align*}
  P(U) = \underset{\vec{r}}{\prod}\, p(u(\vec{r})) &=
  \underset{\vec{r}}{\prod}\, p_0(u(\vec{r})) \, p_1(u(\vec{r})), \\
  p_0(u(\vec{r})) &= \mathcal{N}( u(\vec{r}) \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(u(\vec{r})) &= \mathcal{N}(  \nabla \cdot u(\vec{r}) \mid 0, \mathbf{B}^{-1}),
  \end{align*}

  imposing that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Since the anisotropy is generally aligned with the imaging axes, these will be simplified
  in the following for the sake of clarity:

@@

Finally, the maximum a posteriori problem is adapted into a variational one where we look for
  the minimum of an energy functional, by applying a log-transform:

@@

  this expression is dual to the energy functional corresponding
  to a discrete \gls*{acwe} framework \citep{chan_active_2001}
  with the corresponding anisotropic regularization term \citep{nagel_investigation_1986}.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inner normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	  the vertex $\vec{v}_i$ in each limiting region (see \nameref{app:shape_priors}).
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	  box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer.
	Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	  the corresponding weights $\psi_k(\vec{v}_i)$ \eqref{eq:basis_derivative}.
	}\label{fig:method}
\end{figure}

@@

We can apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to the appendix \nameref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

@@

  in this case, the formulation has been adapted to the non-binary case, $\{l,m\}$
    being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
    them such that
    $\vec{x}' = \vec{v}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$
    and $\bar{s}_i'$ is the speed vector projected on to the unit inward normal to the contour
    at $c_i'$ (described in \autoref{fig:method}).

@@

\paragraph*{Settings, implementation details, and convergence}
\label{sec:conv_report}
The registration parameters (such as $\tau$, the B-Spline grid resolutions,
  target image smoothing, etc.) and implementation details (for instance the sparse matrix approach
  to fast interpolation) are dicussed in the \suppl{}, section S1.
The actual choices of parameter settings are publicly distributed with the source code of the experiments.
Additionally, we release our software along with a tool for generating convergence reports to
  demonstrate the behavior of \emph{regseg} and help scientists configure their own experiments.
One sample report is found in the \suppl{}, section S1.3.

\subsection*{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
In order to demonstrate the performance of \emph{regseg}, we first conducted a battery of
  accuarcy tests over synthetic data, and then applied it in the susceptibility distortion
  correction of real data.
\autoref{fig:evworkflows} presents the workflow implementing the evaluation instruments.
Besides a visual assessment of the results, we report quantitative evaluations using
  two metrics.
In the case of the phantoms, since we had produced distortions along the three
  available dimensions, we computed the Haussdorf distance, by reusing the
  ``point-to-cell'' method of \cite{commandeur_vtk_2011}.
Conversely, the susceptibility-derived distortions only happen along the \gls*{pe}
  axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $w_i$:

  \begin{equation}
  sWI = \frac{1}{P} \sum\limits_p^P \frac{1}{A_p} \sum\limits_i^{N_p} w_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \label{eq:swindex}
  \end{equation}

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figures/figure02}
\caption{Experimental workflow applied on real data from the \acrfull*{hcp}.
  1) The prior surfaces are extracted from the anatomical reference (\gls*{t1} image).
	2) To operate as ground truth, we generate a plausible-synthetic distortion $U_{true}$
	  from the fieldmap using \eqref{eq:fieldmap}.
	3) The \gls*{dmri} data are warped using $U^{-1}_{true}$ to reproduce the effects of real
	  susceptibility-derived distortions.
	Target diffusion scalars (\gls*{fa} and \gls*{adc}) are computed on the distorted data and
		stacked to feed the multivariate input required by our algorithm.
	4) \emph{Regseg} is run, obtaining a $U_{test} = \hat{U}_{true}$, the estimation of
	  the ground-truth deformation.
	  A cross-comparison methodology is also applied, to obtain a competing $\hat{U}_{cc}$.
	5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure*}

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each phantom type, with two resolutions each
  ($1.0mm$ and $2.0mm$ isotropic).
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted using the marching-cubes algorithm
  shipped with \emph{FreeSurfer} \citep{fischl_freesurfer_2012}
  at the higher resolution (1.0mm isotropic).

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus, $\Gamma_{Tha}$;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures, $\Gamma_{VdGM}$;
  3) cerebral \gls*{wm}, $\Gamma_{WM}$;
  4) brain stem and cerebellar \gls*{wm}, $\Gamma_{Bst}$;
	5) cerebellar \gls*{gm}, $\Gamma_{cbGM}$; and
	6) cortical \gls*{gm} surface, $\Gamma_{pial}$.
The choice of this particular model is further addressed in \nameref{sec:res_model_and_metric}.

In this case, we derived the deformation $U_{true}$ from the field maps released with
  the corresponding packages of each dataset from the \gls*{hcp} applying \eqref{eq:fieldmap},
  mimicking the real distortions, using a derivation of our previous work
  \citep{esteban_simulationbased_2014}.
After warping the original \gls*{dmri} with $U_{true}^{-1}$, we computed the \gls*{dti} and
  the derived scalars (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.
